---
title: "Co-clustering Ordinal Data: Simulated and Quality-of-Life Examples"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Co-clustering Ordinal Data: Simulated and Quality-of-Life Examples}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

## 1. Introduction

Standard clustering only groups rows while co-clustering group both rows and columns simultaneously, reducing the dataset into a concise summary of "types of observations" and "types of features" within homogeneous blocks of data highlighting the interactions between them.

The `CoCluster` package allows for the co-clustering of an $N \times D$ data matrix $\mathbf{x}$. Specifically, it groups rows and columns at the same time into $K$ and $L$ row and column clusters with prior probability $\alpha_k$ and $\beta_l$ respectively.

The package implements the **Latent Block Model (LBM)**, which assumes:

- there exists an underlying partition of rows into $K$ clusters (where $\mathbf{v}_{ik}=1$ if row $i$ is in cluster $k$) and columns into $L$ clusters (where $\mathbf{w}_{jl}=1$ if column $j$ is in cluster $l$).

- conditional on the row and column assignments ($v$ and $w$), the observations $\mathbf{x}_{ij}$ are independent and drawn from a probability distribution $P(.;\pi_{kl})$ specific to the block $(v_{ik}=1, w_{jl}=1)$. The fundamental joint probability of the data is given by:

$$
P(\mathbf{x}|\theta) = \sum_{v \in V} \sum_{w \in W} \prod_{ik} \alpha_{k}^{v_{ik}} \prod_{jl} \beta_{l}^{w_{jl}} \prod_{ijkl} P(\mathbf{x}_{ij}; \pi_{kl})^{v_{ik}w_{jl}}
$$

Due to the log likelihood being not tractable because of the sums, the **Stochastic Expectation Maximization (SEM)** algorithm is used which repeats the 2 steps:

- SE step: use the **Gibbs Sampler** to alternatively sample $v|\mathbf{x},w,\theta$ and $v|\mathbf{x},w,\theta$ from their distributions.
- M step: Maximize expected complete log-likelihood: $\theta=\arg \max_{\theta} E(\mathcal{L}_c(\mathbf{x}, v, w,\theta)|\mathbf{x},\theta)$ which can be done using the **Expectation Maximization (EM)** algorithm to estimate the block parameters.

After a burn-in period, $\hat{v}, \hat{w}$ can be obtained by sampling using $\hat{\theta}$ and can used for clustering.

In many domains like psychometrics or surveys, data is ordinal (e.g., Likert scales from 1 to 5). This package natively equips the LBM with the **CUB (Combination of Uniform and shifted Binomial)** observation model for ordinal data. An ordinal response $R \in \{1, \dots, m\}$ is modeled as a mixture of a deliberate choice (shifted binomial, parameterized by a feeling parameter $\xi$) and a random choice (discrete uniform distribution, representing uncertainty), governed by a mixing weight $\pi$:

$$P(\mathbf{x}_{ij} = r \mid v_{ik}=1, w_{jl}=1) = \pi_{kl} \binom{m-1}{r-1} (1-\xi_{kl})^{r-1} \xi_{kl}^{m-r} + \frac{1-\pi_{kl}}{m}$$

A value of $\xi_{kl}$ close to 0 corresponds to a tendency for higher ordinal ratings in that block, while $\pi_{kl}$ near 1 means less random guessing.

**Selecting K and L** The optimal number of row clusters ($K$) and column clusters ($L$) is selected using the **Integrated Completed Likelihood (ICL)** criterion, which penalizes the log-likelihood for model complexity (number of parameters):

$$\text{ICL} = \mathcal{L}_c(\mathbf{x}, v, w,\theta) - \frac{K-1}{2}\log N - \frac{L-1}{2}\log D - \frac{KLv}{2}\log(ND) $$

where $v$ is the number of parameters per block (2 for the CUB model).

## 2. Installation and Setup

You can install the package directly from the source archive. Once installed, load it into your R session:

```{r setup}
library(CoCluster)
```

## 3. Basic Usage: Co-clustering Simulated Data

For a simple use case, we demonstrate how to use the package on a generic dataset. We will use the `Msimulated` dataset provided by the `ordinalClust` package.

```{r}
data(Msimulated, package = "ordinalClust")
x_sim <- as.matrix(Msimulated)
m_sim <- max(x_sim, na.rm = TRUE) # Detect the number of ordinal categories
```

### Fitting the Model

The main function of the package is `cocluster()`. We fit an LBM with the CUB observation model on the simulated data, enabling `choose_n_clusters = TRUE` so the algorithm automatically searches the neighborhood of our starting $K$ and $L$ to greedily find the optimal dimensions based on ICL.

```{r, results="hide"}
result_sim <- cocluster(
  x = x_sim,
  K = 3,              
  L = 3,              
  model = cub(m_sim),
  n_init = 4,
  n_iterations = 40,
  n_gibbs_iterations = 2,
  choose_n_clusters = TRUE
)
```

### Interpretation of Results

**Model Selection:** Once the model is fitted, the `cocluster` object contains the best model selected by ICL under `$selected_lbm`.

```{r}
best_sim <- result_sim$selected_lbm
print(c(K_optimal = best_sim$K$row, L_optimal = best_sim$K$col))
```

This tells us the optimal number of clusters discovered by the ICL greedy search. We can also inspect the ICL values of all the models evaluated during the search:

```{r}
data.frame(
  K = sapply(result_sim$lbms, function(m) m$K$row),
  L = sapply(result_sim$lbms, function(m) m$K$col),
  ICL = sapply(result_sim$lbms, function(m) round(m$icl, 2))
)
```

**Extracting Clusters:** To obtain the hard cluster assignments for each row and column, you can inspect $z$:

```{r}
table(best_sim$z$col)
```

**Cluster Proportions:** You can retrieve the estimated mixing proportions for the rows and columns:

```{r}
print(best_sim$p$row)
print(best_sim$p$col)
```

These vectors represent the estimated prior proportions of the clusters. For example, the values in `p$row` indicate the expected fraction of the total rows belonging to each row cluster.

**Visualization:** We can visually verify the clustering structure by sorting the matrix by the predicted row and column clusters:

```{r, fig.width=5, fig.height=5}
sorted_x_sim <- x_sim[order(best_sim$z$row), order(best_sim$z$col)]
image(x = 1:ncol(sorted_x_sim), y = 1:nrow(sorted_x_sim),
      z = t(sorted_x_sim)[, nrow(sorted_x_sim):1],
      col = heat.colors(m_sim), axes = FALSE,
      xlab = "Column Clusters", ylab = "Row Clusters",
      main = "Simulated Co-clustered Data Matrix")

# Label the clusters exactly at their centers
axis(1, at = tapply(1:ncol(sorted_x_sim), sort(best_sim$z$col), mean), labels = paste("Cluster", 1:best_sim$K$col))
axis(2, at = tapply(nrow(sorted_x_sim):1, sort(best_sim$z$row), mean), labels = paste("Cluster", 1:best_sim$K$row))

# Draw cluster boundaries
abline(v = cumsum(table(best_sim$z$col)) + 0.5, lwd = 2)
abline(h = nrow(sorted_x_sim) - cumsum(table(best_sim$z$row)) + 0.5, lwd = 2)
box()
```

## 4. Advanced Usage: Quality-of-Life Questionnaire Data

For a more advanced use case, we will co-cluster a real-world dataset and explore the package's hyper-parameters and visualization capabilities.

### Description of the Dataset

We use the `dataqol` dataset from the `ordinalClust` package. It consists of a quality-of-life questionnaire containing patients (rows) and 30 questions (columns).

The CUB observation model we are using requires that all ordinal variables share the exact same number of categories $m$. In the `dataqol` dataset, questions 1 through 28 have 4 categories, while questions 29 and 30 have 7 categories. Therefore, we must restrict our dataset to `q1` through `q28` and set $m=4$. We also drop the patient ID column.

```{r data_qol}
data(dataqol, package = "ordinalClust")
x_qol <- as.matrix(dataqol[, paste0("q", 1:28)])
m_qol <- 4

table(x_qol, useNA = "ifany")
```

### Advanced Model Fitting

We can finely tune the SEM-Gibbs algorithm using several optional parameters: `n_init`: The number of random initializations to avoid local maxima, `n_iterations`: Total number of SEM iterations, `n_gibbs_iterations`: The number of Gibbs sampling passes inside the E-step, `eps`: Convergence threshold for the log-likelihood, `burn_in`: The number of iterations to discard before averaging the parameter estimates, etc, see `?cocluster`.

```{r, results="hide"}
set.seed(1)
result_qol <- cocluster(
  x = x_qol,
  K = 3, L = 3,
  model = cub(m_qol),
  n_init = 10,
  n_iterations = 100,
  n_gibbs_iterations = 4,
  eps = 1e-6,
  burn_in = 40,
  choose_n_clusters = FALSE
)
best_qol <- result_qol$selected_lbm
```

### Model Selection

Since we disabled the greedy search (`choose_n_clusters = FALSE`) for execution speed in this vignette, the algorithm directly returns the model for our specified $K=3$ and $L=3$.

```{r}
print(c(K = best_qol$K$row, L = best_qol$K$col, ICL = round(best_qol$icl, 2)))
```

### Extracting Clusters and Proportions

We can extract the estimated hard cluster assignments and their prior proportions:

```{r}
# Sizes of the estimated column clusters
table(best_qol$z$col)

# Estimated prior proportions of the row and column clusters
print(best_qol$p$row)
print(best_qol$p$col)
```

### Interpreting Block Parameters

The estimated CUB parameters for each block (`pi` is the weight of the deliberate choice, `xi` is the feeling parameter) are stored in `block_parameters`:

```{r blocks}
blocks <- best_qol$block_parameters
data.frame(
  row_cluster = rep(seq_len(nrow(blocks)), ncol(blocks)),
  col_cluster = rep(seq_len(ncol(blocks)), each = nrow(blocks)),
  pi = sapply(blocks, function(b) round(b$pi_, 3)),
  xi = sapply(blocks, function(b) round(b$xi, 3))
)
```

**How to interpret:** A block with a low `xi` contains mostly high ratings, while a block with a high `xi` contains mostly low ratings. Additionally, a `pi` near 1 indicates that the responses in that block are deliberate and consistent, whereas a lower `pi` suggests more randomness and uncertainty in the ratings.

### Visualization

To visualize the clusters, you can sort your data matrix by the estimated row and column assignments and plot it:

```{r, fig.width=5, fig.height=5}
# Sort data by cluster assignments
sorted_x <- x_qol[order(best_qol$z$row), order(best_qol$z$col)]

# Visualize the clustered blocks
image(x = 1:ncol(sorted_x), y = 1:nrow(sorted_x),
      z = t(sorted_x)[, nrow(sorted_x):1],
      col = heat.colors(m_qol), axes = FALSE,
      xlab = "Question Clusters", ylab = "Patient Clusters",
      main = "Co-clustered Data Matrix")

# Label the clusters exactly at their centers
axis(1, at = tapply(1:ncol(sorted_x), sort(best_qol$z$col), mean), labels = paste("Cluster", 1:best_qol$K$col))
axis(2, at = tapply(nrow(sorted_x):1, sort(best_qol$z$row), mean), labels = paste("Cluster", 1:best_qol$K$row))

# Draw cluster boundaries
abline(v = cumsum(table(best_qol$z$col)) + 0.5, lwd = 2)
abline(h = nrow(sorted_x) - cumsum(table(best_qol$z$row)) + 0.5, lwd = 2)
box()
```

### Grid Search Model Selection

If you prefer to perform an exhaustive grid search instead of using the local greedy search (`choose_n_clusters = TRUE`), you can simply pass a vector of values to the `K` and `L` arguments in `cocluster()`. The function will evaluate every combination and return the best overall model.

```{r grid_search, eval = FALSE}
result_grid <- cocluster(
  x = x_qol,
  K = 2:4, # Test K from 2 to 4
  L = 2:4, # Test L from 2 to 4
  model = cub(m_qol),
  n_init = 3,
  n_iterations = 10,
  n_gibbs_iterations = 2,
  eps = 1e-6,
  burn_in = 5
)

# View the ICL of all models tested in the 3x3 grid
data.frame(
  K = sapply(result_grid$lbms, function(m) m$K$row),
  L = sapply(result_grid$lbms, function(m) m$K$col),
  ICL = sapply(result_grid$lbms, function(m) round(m$icl, 2))
)
```

*(This code block is not evaluated here as an exhaustive grid search takes a couple of minutes to run.)*

## 5. Using Different Observation Models

The package is designed to be extensible. You are not strictly limited to the CUB model. The `model` argument in `cocluster()` accepts any list of functions defining an observation model. To implement your own model (e.g., a Poisson or Gaussian block model), you simply need to provide a list containing: 1. `probability(...)`: Computes the probability of a single observation. 2. `log_probability(...)`: Computes the log-probability of an entire matrix block. 3. `estimate_parameters(...)`: Returns a list of estimated parameters for a given block matrix. 4. `number_of_parameters`: An integer used to compute the ICL penalty. For more information, see `?cocluster`.

## 6. References

<https://inria.hal.science/hal-01658589/document>

<https://inria.hal.science/inria-00494796/document>

Jacques and Biernacki (2018) performed a co-clustering of this exact `dataqol` ordinal dataset using the BOS model. The number of blocks and the groupings of the questions can be compared directly with our results, keeping in mind that the underlying assumptions of the CUB and BOS models differ.
