---
title: "Coclustering simulated ordinal data"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Coclustering simulated ordinal data}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

## Introduction

This vignette demonstrates how to use the `CoCluster` package to co-cluster ordinal data using a Latent Block Model (LBM) equipped with the CUB (Combination of Uniform and shifted Binomial) observation model.

We will simulate a dataset where observations are ordinal ratings (e.g., from 1 to 5), and then recover the underlying row and column clusters.

```{r setup}
library(CoCluster)
```

## Simulating the Data

First, we define a helper function to simulate data from the CUB model. The CUB model assumes that a rating is generated from a mixture of two components:
1. A deliberate choice (shifted binomial distribution).
2. A random choice (discrete uniform distribution).

```{r}
# Simulate 'n' responses from a CUB model with 'm' categories
simulate_cub <- function(n, m, pi_, xi) {
  # Component 1: shifted binomial
  deliberate <- rbinom(n, size = m - 1, prob = 1 - xi) + 1
  
  # Component 2: uniform
  random <- sample(1:m, size = n, replace = TRUE)
  
  # Mixture indicator: 1 if deliberate, 0 if random
  is_deliberate <- rbinom(n, size = 1, prob = pi_) == 1
  
  # Final response
  ifelse(is_deliberate, deliberate, random)
}
```

Now we simulate a dataset of size $100 \times 80$ with $K=3$ row clusters and $L=2$ column clusters.

```{r}
set.seed(42)

N <- 100
D <- 80
K_true <- 3
L_true <- 2
m <- 5 # Categories: 1, 2, 3, 4, 5

# Cluster proportions
p_row <- c(0.4, 0.3, 0.3)
p_col <- c(0.6, 0.4)

# True cluster assignments
z_row <- sample(1:K_true, size = N, replace = TRUE, prob = p_row)
z_col <- sample(1:L_true, size = D, replace = TRUE, prob = p_col)

# Define pi and xi parameters for each of the 3x2 blocks
pi_true <- matrix(c(
  0.9, 0.8, # Row cluster 1
  0.2, 0.9, # Row cluster 2
  0.8, 0.8  # Row cluster 3
), nrow = K_true, ncol = L_true, byrow = TRUE)

xi_true <- matrix(c(
  0.1, 0.9, # Row cluster 1: High ratings, Low ratings
  0.5, 0.5, # Row cluster 2: Medium ratings (but left is mostly uniform due to low pi)
  0.8, 0.2  # Row cluster 3: Low ratings, High ratings
), nrow = K_true, ncol = L_true, byrow = TRUE)

# Generate the data matrix
x <- matrix(NA, nrow = N, ncol = D)
for (i in 1:N) {
  for (j in 1:D) {
    k <- z_row[i]
    l <- z_col[j]
    x[i, j] <- simulate_cub(1, m, pi_true[k, l], xi_true[k, l])
  }
}
```

## Fitting the Model

We can now use the `cocluster()` function to fit an LBM with the `cub(m)` observation model. 
Because SEM-Gibbs relies on random initializations to avoid local maxima, we use `n_init = 5` for speed in this vignette (though higher values like 20 are recommended for real data). We set `choose_n_clusters = FALSE` and provide the true number of clusters to skip the greedy model search for this demonstration.

```{r, results="hide"}
# Fit the model
result <- cocluster(
  x = x,
  K = 3,
  L = 2,
  model = cub(m),
  n_init = 5,
  n_iterations = 100,
  n_gibbs_iterations = 5,
  choose_n_clusters = FALSE
)
```

## Extracting Results

Let's inspect the results from the selected LBM model.

```{r}
best_model <- result$selected_lbm

# Estimated row cluster proportions
print(best_model$p$row)

# Estimated column cluster proportions
print(best_model$p$col)
```

We can also compare the estimated block parameters against our true parameters.

```{r}
# Block 1,1 Estimated Parameters
print(best_model$block_parameters[[1, 1]])

# Cross-tabulate our true cluster assignments with the estimated ones 
# (Note: Cluster labels might be permuted compared to the true labels)
table(True = z_row, Estimated = best_model$z$row)
table(True = z_col, Estimated = best_model$z$col)
```
